Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-09-13 manuscript/manuscript.tex
requested text changes, line breaks, comments removed (#136)

\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}

@@

In this tutorial, we assume the reader's familiarity with the basic theory behind the MSM approach (see Sec.~\ref{sec:background}) and focus on usage of PyEMMA.
Nevertheless, we will mention important theoretical concepts when appropriate throughout the tutorial.
We also assume the reader is familiar with MD analyses of proteins and peptides and commonly used structural components of these systems (for a review of the simulation of biomolecular macromolecules, see~\cite{dror2012biomolecular}).

@@

\subsection{Essential theory}
\label{sec:theory}

Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states that adhere to the Markov property.
The Markov property asserts that the dynamics in the state space is ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ after a time $\tau$ is independent of the history of the system before it was in $i$.
%This lag time must be sufficiently short to resolve the dynamics of interest, but long enough such that the Markovian approximation is appropriate.

In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
Then, the count matrix is converted to a row-stochastic transition probability matrix, which is defined for the specified lag time.
To ensure that the transition probability matrix has desirable mathematical properties, detailed balance is enforced when converting the count matrix to the transition probability matrix, which requires that,

\begin{equation}
\label{eq:balance}
p_t(i) p(i \rightarrow j) = p_t(j)p(j \rightarrow i),
\end{equation}

\noindent{}where $p_t(i)$ represents the probability of being in state $i$ at time $t$, and $p(i \rightarrow j)$ is the probability of transitioning from state $i$ to $j$ at the next time step.
% Detailed balance requires that the probability of transitioning from state $i$ to state $j$, conditioned upon the system being in state $i$, is the same as the probability of transitioning from state $j$ to state $i$ conditioned upon the system being in state $j$.
The requirement of detailed balance indicates that we assume our system is reversible.
We additionally require that the system is ergodic, which means that every state is accessible from every other state.

When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our state space, but short enough to resolve the dynamics in which we are interested.
Plotting the implied timescales (ITS) as a function of~$\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via,
\begin{equation}
\label{eq:its}
t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
\end{equation}

\noindent{}When the ITS become approximately constant with the lag time, we say that our timescales have converged and choose the smallest lag time with the converged timescales in order to maximize the model's temporal resolution.

Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{msm-jhp}.
The CK property for a Markovian matrix is,

\begin{equation}
T(k \tau) = T^k(\tau),
\end{equation}

\noindent{}where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.

Once validated, the transition matrix can be decomposed into left eigenvectors and eigenvalues,

\begin{equation}
\label{eq:spectral_left}
\phi_i^\top  T(\tau) = \lambda_i(\tau) \phi_i^\top,
\end{equation}

\noindent{}or equivalently into their right eigenvectors and eigenvalues,

\begin{equation}
\label{eq:spectral_right}
T(\tau)\psi_i = \lambda_i(\tau) \psi_i,
\end{equation}

\noindent{}where the eigenvalue-eigenvector pairs are indexed in decreasing order. 
The eigenvalues are the same in both cases, however, the left and right eigenvectors are related to each other as

\begin{equation}
\phi_i = \pi \circ \psi_i,
\label{eq:left-right-eigenvalue-relation}
\end{equation}
\noindent{}where $\pi$ is the \emph{stationary distribution} of the MSM, and $\circ$ corresponds to an element-wise vector product.

The highest eigenvalue, $\lambda_1(\tau)$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution, $\pi$.
From the relationship between the left and right eigenvectors (eq.~\ref{eq:left-right-eigenvalue-relation}) we see that the right eigenvector $\psi_1$ is a vector consisting of~$1$'s.

The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} timescales of dynamical processes within the system (eq.~\ref{eq:its}).

The right eigenvectors $\psi_i$ each encode a dynamical process (for $i>1$), corresponding to the characteristic time-scale, $t_i$.
The coefficients of the eigenvectors represent the flux into and out of the Markov states that characterizes that process.
Again, as can be seen from (eq.~\ref{eq:left-right-eigenvalue-relation}) the corresponding left eigenvectors $\phi_i$ contain the same information weighted by the stationary distribution.

\subsection{MSM construction the variational approach}
\label{sec:construction}

@@

\subsection{Background knowledge and resources}
\label{sec:background}

\subsection{Software/system requirements}

@@

\section{Content and links}

@@

\subsection{Discretization}
TICA yields a representation of our molecular simulation data with a reduced dimensionality, which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation. Here, we use the $k$-means algorithm to segment the four dimensional TICA space into $k=75$ cluster centers. The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above, which is shown in the showcase notebook (00).

\subsection{MSM estimation and validation}
% When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
% Plotting the implied timescales (ITS) as a function of~$\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}. The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
% \begin{equation}
% \label{eq:its}
% t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
% \end{equation}
A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty. The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with~$100$ samples.
In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above~$0.5$~ns (Fig.~\ref{fig:its-and-ck}a). Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5$~ns.

To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test.
% The CK test compares the right and the left side of the Chapman-Kolmogorov equation
% \begin{equation}
% \label{eq:ck}
% T(k \tau) = T^k(\tau)
% \end{equation}
% where $T$ is the MSM transition matrix. The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a smaller number of metastable states before performing the test.
An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
For this analysis, we chose five metastable states.
The CK test (Fig.~\ref{fig:its-and-ck}b) shows that predictions from our MSM (blue-dashed lines) agrees well with MSMs estimated with longer lag times (black-solid lines)
Thus, the CK test confirms that five metastable states is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (blue/shaded area).

@@

\subsection{Modeling large systems}
When estimating MSMs for large systems, challenges may arise that are mostly system dependent.

A case in point is the curse of dimensionality:~it is difficult to discretize a high dimensional feature space. While it is somewhat computationally demanding, more importantly, Euclidean distances become less meaningful with increasing dimensionality~\cite{aggarwal_surprising_2001} and thus cluster assignments based on that norm may yield a poor discretization.
Especially for large systems, it is thus particularly important to first find a suitable set of features, and to further apply dimensionality reduction techniques (e.g.~TICA, VAMP, if applicable) to obtain a low dimensional representation of the slow dynamics.
Hidden Markov models (HMMs) might further mitigate poor discretization to a certain extent~\cite{noe-proj-hid-msm}.

@@

The present tutorial presents the basics of modern Markov state modeling with PyEMMA. 
However, recent years have seen many extensions of the methodology --- many of which are available within PyEMMA. 
We encourage interested readers to look into these methods in the software documentation and to make use of the specific Jupyter notebooks distributed with PyEMMA.

>>>>

\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}
\usepackage{bm}

@@

In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances. (see [7] for a review).

We further assume that the reader is familiar with the basic ideas and theory underlying Markov modeling and will only give a brief reminder of the basic concepts in Section 2.

@@

\subsection{Markov state models}
\label{sec:theory}

Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states,
and a transition matrix $P(\tau) = [p_{ij}(\tau)]$ denoting the conditional probability of finding the system in state $j$ at time $t+\tau$ given that it was in state $i$ at time $t$.
Let us make two remarks to avoid common misconceptions:
\begin{enumerate}
\item Equilibrium:
While most analysis techniques require simulation trajectories to be long enough to sample from the equilibrium distribution, this is not required for MSMs.
Because MSMs are using the \emph{conditional} probability $p_{ij}(\tau)$,
they are useful to analyze short simulation trajectories with arbitrary starting points---see~\cite{oom-feliks} for restrictions.
\item Markovianity:
An MSM is a memoryless model.
Early MSM papers have argued that accurate MSMs can be found if a few states with high barriers are captured by the MSM states so as to achieve a Mori-Zwanzig projection with fast-decaying memory~\cite{swope-its,noe2007jcp,chodera2007jcp}.
The modern view, however, is that MSMs can be highly accurate if the MSM states discretize the collective coordinates of the slowest processes well~\cite{msm-jhp}.
This mainly requires that the system is characterized by only a few slow processes at lag time $\tau$,
which is true for cooperative systems such as most proteins, but not for highly frustrated systems such as glasses.
\end{enumerate}

In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
Then, the count matrix is converted to a row-stochastic transition probability matrix $P$, which is defined for the specified lag time.
For MD simulations in equilibrium, $P$ should obey detailed balance which is enforced by constraining the estimation of $P$ to the following equations:
\begin{equation}
\label{eq:balance}
\pi_i P_{ij} = \pi_j P_{ji},
\end{equation}
where $\pi_i$ is the stationary probability of state $i$ and $P_{ij}$ is the probability of transitioning to state $j$ conditional on being in state $i$.
The constraints (\ref{eq:balance}) are omitted if MD simulations are not conducted in equilibrium, e.g.,
for systems experiencing a pulling force or an external potential---see~\cite{Koltai2018} for a recent review on nonequilibrium MSMs.
For the remainder of this section we will simplify the matter by assuming the more common scenario of MD simulations without external forces and (\ref{eq:balance}) to hold.

When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our state space, but short enough to resolve the dynamics in which we are interested.
Plotting the implied timescales (ITS) as a function of~$\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via,
\begin{equation}
\label{eq:its}
t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
\end{equation}
When the ITS become approximately constant with the lag time, we say that our timescales have converged and choose the smallest lag time with the converged timescales in order to maximize the model's temporal resolution.

Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways}.
The CK property for a Markovian matrix is,
\begin{equation}
T(k \tau) = T^k(\tau),
\end{equation}
where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.

Once validated, the transition matrix can be decomposed into eigenvectors and eigenvalues.
The highest eigenvalue, $\lambda_1(\tau)$, is unique and equal to $1$.
Its corresponding left eigenvector is the stationary distribution, $\bm{\pi}$:
\begin{equation}
\bm{\pi}^\top  P(\tau) = \bm{\pi}^\top.
\end{equation}

The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} timescales of dynamical processes within the system (eq.~\ref{eq:its}).
The dynamical process themself (for $i>1$) are encoded by the right eigenvectors $\bm{\psi}_i$,
\begin{equation}
P(\tau)\bm{\psi}_i = \lambda_i(\tau) \bm{\psi}_i,
\end{equation}
where the eigenvalue-eigenvector pairs are indexed in decreasing order.
The coefficients of the eigenvectors represent the flux into and out of the Markov states that characterize the corresponding process.
The right eigenvector $\bm{\psi}_1$ is a vector consisting of~$1$'s.

\subsection{Variational approach and TICA}

@@

\subsection{Software and installation}

@@

\section{PyEMMA tutorials}

@@

\subsection{Discretization}

TICA yields a representation of our molecular simulation data with a reduced dimensionality,
which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation.
Here, we use the $k$-means algorithm to segment the four dimensional TICA space into $k=75$ cluster centers.
The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above,
which is shown in the showcase notebook (00).

\subsection{MSM estimation and validation}

A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$;
accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty.
The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with~$100$ samples.
In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above~$0.5$~ns (Fig.~\ref{fig:its-and-ck}a).
Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5$~ns.

To test the validity of our MSM, we perform a Chapman-Kolmogorov (CK) test.
Visualizing the full transition probability matrix $T$ is difficult;
we therefore coarse-grain $T$ into a smaller number of metastable states before performing the test.
An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
For this analysis, we chose five metastable states.
The CK test (Fig.~\ref{fig:its-and-ck}b) shows that predictions from our MSM (blue-dashed lines)
agrees well with MSMs estimated with longer lag times (black-solid lines)
Thus, the CK test confirms that five metastable states is an appropriate choice
and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the
long-timescale behavior of our system within error (blue/shaded area).

@@

\subsection{Modeling large systems}

When estimating MSMs for large systems, challenges may arise that are mostly system dependent.

A case in point is the curse of dimensionality:~it is difficult to discretize a high dimensional feature space.
While it is somewhat computationally demanding, more importantly,
Euclidean distances become less meaningful with increasing dimensionality~\cite{aggarwal_surprising_2001}
and thus cluster assignments based on that norm may yield a poor discretization.
Especially for large systems, it is thus particularly important to first find a suitable set of features,
and to further apply dimensionality reduction techniques (e.g., TICA, VAMP, if applicable)
to obtain a low dimensional representation of the slow dynamics.
Hidden Markov models (HMMs) might further mitigate poor discretization to a certain extent~\cite{noe-proj-hid-msm}.

@@

The present tutorial presents the basics of modern Markov state modeling with PyEMMA. 
However, recent years have seen many extensions of the methodology---many of which are available within PyEMMA. 
We encourage interested readers to look into these methods in the software documentation and to make use of the specific Jupyter notebooks distributed with PyEMMA.
